%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plantilla TFG/TFM
% Universidad de Murcia. Facultad de Informática
% Realizado por: José Manuel Requena Plens
% Modificado: Pablo José Rocamora Zamora
% Contacto: pablojoserocamora@gmail.com
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\chapter*{Extended Abstract}
\addcontentsline{toc}{chapter}{Extended Abstract}

Cryo-Electron Tomography (cryo-ET) is a structural biology imaging technique that enables the
three-dimensional reconstruction of biological specimens—such as individual proteins, protein
complexes, and intact cells—at near-native conditions and nanometric resolution. Unlike classical
structural methods such as X-ray crystallography or Nuclear Magnetic Resonance, cryo-ET does not
require sample crystallisation or chemical fixation; instead, specimens are rapidly vitrified in
liquid ethane to prevent ice crystal formation, preserving their native hydrated structure. During
acquisition, the electron beam traverses the sample at a series of tilt angles, typically ranging
from $-60°$ to $+60°$, and the resulting 2D projection images are computationally back-projected
and filtered to reconstruct a 3D density map known as a tomogram. Each voxel in the tomogram
encodes a local scattering density, which correlates with the distribution of atomic mass in the
corresponding region of the sample. Cryo-ET has become an indispensable tool in structural cell
biology for studying macromolecular machines, membrane architecture, and organelle organisation
in their cellular context.

The automated analysis of cryo-ET data has benefited enormously from deep learning methods over
the past decade. Convolutional neural networks and related architectures have been applied to tasks
such as tomogram denoising, organelle segmentation, sub-tomogram classification, and particle
picking. These supervised approaches, however, require large annotated training datasets: volumes
in which each voxel or region is labelled with its biological identity. Generating such ground-truth
labels experimentally is extraordinarily tedious and relies on expert manual curation, which
severely limits the scale at which labelled data can be produced. This data scarcity problem is
one of the primary bottlenecks in the application of deep learning to cryo-ET.

A well-established strategy to overcome this limitation is the generation of \textbf{synthetic
tomograms}: computationally produced volumes that simulate the physical image-formation process
of the electron microscope, including the projection geometry, the contrast transfer function,
missing-wedge artefacts arising from the limited tilt range, and realistic noise models.
When synthetic data is sufficiently faithful to experimental observations, networks trained on it
can generalise to real tomograms, reducing the need for manual annotation. The fidelity of
synthetic data depends critically on how realistically it reproduces the spatial distribution
and density of macromolecules within the simulation volume.

\textit{PolNet} is an open-source Python framework specifically designed for the generation of
synthetic cryo-ET datasets. It provides a modular simulation pipeline that models lipid membranes,
transmembrane protein complexes, and cytoplasmic macromolecular assemblies, ultimately producing
synthetic tomograms together with their ground-truth segmentation masks. The protein insertion
module of PolNet is based on \textit{SAWLC} (Surface Aware Walk from Last Coordinate), a stochastic
placement strategy in which each new protein is seeded near the surface of the most recently
inserted molecule. This surface-aware approach produces spatially clustered distributions that
resemble the crowded, heterogeneous organisation observed in real cells. Despite its biological
motivation, SAWLC is computationally demanding: the sequential nature of the placement process,
combined with high rejection rates in dense volumes, means that generating a single tomogram with
thousands of macromolecules can require several hours of CPU computation. This throughput limitation
restricts the scale at which synthetic datasets can be produced and motivates the investigation of
more efficient protein insertion strategies.

The present Bachelor's Thesis addresses this bottleneck by designing, implementing, and
benchmarking \textbf{Tetris~3D}, a novel protein insertion algorithm that uses three-dimensional
FFT-based cross-correlation to determine, at each step, the optimal position for placing a new
macromolecule within the simulation volume. The algorithm is developed and evaluated in an
independent repository with the explicit goal of comparing its performance against SAWLC and
assessing whether it constitutes a viable replacement for the insertion module of PolNet.

The core idea behind Tetris~3D is to reframe protein placement as an optimisation problem over
the space of all geometrically valid positions within a local search region. Given the current
state of the output volume and a new protein density map, the algorithm first extracts a local
bounding box around a randomly sampled seed coordinate. Within this bounding box, a binary
occupancy map is constructed by thresholding the existing density at an isosurface value derived
from the protein's own maximum intensity. A shell template is then generated by morphologically
dilating the binary protein mask and subtracting the protein itself: the resulting shell encodes
the ideal contact zone where a new protein should be placed to fit tightly against its neighbours
without overlapping. The three-dimensional cross-correlation between this shell template and the
binary occupancy map is computed using the Fast Fourier Transform (FFT), producing a score map
in which high values indicate positions where the new protein would make close contact with
existing structure. The position with the maximum valid score—subject to boundary constraints
and strict non-overlap verification—is selected, and the protein density is added to the output
volume at that location.

This formulation has two important consequences. First, each insertion finds the \textit{globally
best} valid position within the search region, rather than accepting the first non-overlapping
candidate discovered by random sampling. This leads to denser packing: proteins are placed in
a manner that minimises wasted interstitial space, achieving higher volumetric occupancy at the
end of the simulation. Second, the FFT-based score computation evaluates all candidate positions
simultaneously in $\mathcal{O}(N \log N)$ time, where $N$ is the number of voxels in the
bounding box, rather than requiring a separate overlap check for each candidate. This is
fundamentally more efficient than the iterative random-walk approach of SAWLC.

A GPU-accelerated variant of Tetris~3D was also developed using the \textit{CuPy} library,
which provides a NumPy-compatible API for array operations on NVIDIA graphics processing units.
The straightforward approach of porting the algorithm to CuPy is not sufficient to achieve
optimal GPU utilisation, however, because the native overlap validation loop requires repeated
synchronisation between the host (CPU) and the device (GPU) to evaluate each candidate position.
Each synchronisation stalls the GPU pipeline and incurs significant latency. To address this,
the GPU implementation adopts a \textbf{two-FFT strategy}: a second cross-correlation is computed
between the binary occupancy map and the binary protein mask, producing an overlap map that
simultaneously encodes, for every position in the bounding box, the degree of voxel overlap
between the candidate placement and the existing structure. By combining the shell-correlation
score map with this overlap map using vectorised boolean masking, all invalid positions are
eliminated in a single GPU kernel without any loop. The position with the maximum remaining
score is then retrieved with a single host-device data transfer per insertion, regardless of
the size of the bounding box. This reduces the number of GPU–CPU synchronisation points from
up to $\mathcal{O}(50)$ per insertion to exactly one, dramatically improving GPU pipeline
throughput and delivering substantial speedup over both the CPU variant and SAWLC.

To evaluate Tetris~3D systematically, a benchmarking suite was designed comprising three
complementary scripts. The first, \texttt{plot\_comparison.py}, runs complete simulations with
all eleven protein types for each of the three algorithm variants (Tetris CPU, Tetris GPU, and
SAWLC) and generates comparative visualisations of volumetric occupancy, cumulative execution
time, monomer population per protein type, and insertion throughput. The second script,
\texttt{profile\_comparison.py}, produces occupancy-vs-time curves that show how quickly each
algorithm fills the available volume over the course of a full simulation. The third,
\texttt{profile\_test.py}, instruments the Tetris algorithm with fine-grained timers to decompose
the total execution time into its constituent phases: seed selection, random rotation, binary
template construction, FFT cross-correlation, and overlap validation. This decomposition
identifies where time is actually spent and guides further optimisation efforts.

All simulations share the same experimental configuration: eleven protein types at 10~Å
resolution, the same membrane volume derived from a real cryo-ET tomogram, and identical
placement parameters including the isosurface threshold ratio and the maximum number of
consecutive failures before moving to the next protein type. Each algorithm runs in an isolated
subprocess spawned with Python's \texttt{multiprocessing} module using the \texttt{spawn}
start method, ensuring a clean import environment. Results are serialised to a JSON cache file
so that visualisations can be regenerated without repeating the time-consuming simulations.

The preliminary results of this comparative study confirm the expected benefits of the Tetris~3D
approach. Across all simulated scenarios, Tetris~3D achieves higher volumetric occupancy than
SAWLC at the end of a full simulation run. The FFT-guided placement strategy packs proteins more
efficiently because it actively selects the contact position that maximises adjacency to existing
structure, whereas SAWLC's random walk may leave isolated free-volume pockets that are too
narrow to accommodate any remaining protein. This difference in packing quality is particularly
pronounced for smaller proteins, which can fill residual interstitial space that larger molecules
cannot access: the Tetris~3D algorithm inserts substantially more monomers of small proteins than
SAWLC, contributing more to the final volumetric occupancy.

In terms of execution time and throughput, the GPU variant of Tetris~3D delivers the best
performance. By eliminating the sequential overlap-checking loop and replacing it with fully
vectorised GPU computation, the two-FFT optimisation achieves near-constant cost per insertion
independent of the current density of the volume. This is in contrast to the CPU and SAWLC
implementations, where the cost per insertion tends to increase as the volume becomes denser
and more candidate positions need to be rejected. As a consequence, the performance advantage of
Tetris~3D GPU over SAWLC grows as the simulation progresses. The CPU variant of Tetris~3D also
outperforms SAWLC in throughput, because the FFT cross-correlation evaluates all positions in
a single pass rather than through repeated random sampling.

It is important to acknowledge that the two algorithms reflect different biological priorities.
SAWLC is designed to reproduce the spatial clustering patterns observed in real cells, where
proteins are not uniformly distributed but tend to aggregate near membrane surfaces or form
oligomeric assemblies. Tetris~3D, in its current form, optimises for volumetric packing
efficiency rather than for biological realism in the spatial distribution. This trade-off should
be considered when choosing between the two algorithms for a given simulation goal: if the
primary objective is to generate dense synthetic tomograms for training detection or segmentation
networks, Tetris~3D is the more appropriate choice; if the goal is to reproduce specific spatial
statistics of macromolecular distributions, SAWLC may better capture those features.

In conclusion, this work demonstrates that Tetris~3D is a computationally efficient alternative
to SAWLC for protein insertion in synthetic cryo-ET tomogram generation. The algorithm achieves
higher packing densities and shorter runtimes through FFT-based global optimisation and
GPU-accelerated vectorised computation, positioning it as a strong candidate for replacing the
SAWLC module in PolNet pipelines where throughput and dataset scale are the primary concerns.
Future work will focus on incorporating biologically motivated spatial constraints into the
algorithm—for example, surface-proximate seeding similar to SAWLC—to improve the realism of
the generated distributions while retaining the computational advantages of the FFT-based
placement strategy. Further directions include extending the algorithm to handle elongated
filamentous assemblies, conducting a larger-scale evaluation across diverse tomogram geometries
and protein libraries, and exploring the use of Tetris~3D as a pre-processing step that
initialises SAWLC simulations with a dense uniform packing before applying surface-aware
refinement.
